Adding selection to \stdpopsim: oh how the tables have turned

### 2024-11-15 analysis2_manuscript.tex
revised sweep section

    ** this section needs revision based on new figures that Murillo will add **
    In Figure \ref{fig:chr1_power} we show the power to detect selective sweeps at 100 locations along human chromosome 1 from
    the simulations described above. We show the results of three methods for detecting sweeps: 1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos}, and 3) an empirical cutoff of $\pi$ to find regions of reduced diversity.
    A few things are immediately apparent from this figure. First, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that the YRI sample has a larger effective population size and thus
    more variation to detect a sweep. Further the population size of YRI has been relatively constant over the last 100,000 years,
    thus the out-of-Africa bottleneck has does not mute the signal of a sweep as much as in the CEU sample. This is consistent with
    what we know about the effects of demography on the power to detect sweeps \citep[e.g.,][]{simonsen1995properties}.
    Second, the power to detect sweeps is slightly lower in the presence of purifying selection acting on exons but the effect is not
    as strong as one might expect. This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Finally and most importantly, the power to detect sweeps varies considerably among regions of the genome. 
    Indeed variation in power along the chromosome is much greater than the variation in power between the CEU and YRI samples or between
    models with or without purifying selection. This is likely due to the effects of recombination rate heterogeniety as well
    as the density of functional elements along the chromosome.

    %sweep power vs recombination rate
    \begin{figure}
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_cM.pdf}
        \caption{
        Power to detect selective sweeps at 100 locations along human chromosome 1 under a three population out-of-Africa model \citep{gutenkunst2009inferring}  
        with and without purifying selection as a function of local recombination rate.
        Simulations are as described in Figure \ref{fig:chr1_power}.
        Each panel shows the results of simulations with purifying selection acting on exons (red; labelled BGS)
        and without purifying selection acting on exons (blue; labelled Neutral).
        True positive rate is shown for these methods for samples from the CEU and YRI samples (left and right respectively).
        Points represent the power at each location and trend lines along with 95\% confidence intervals are shown fit to those points.
        }
        \label{fig:power-recomb}
    \end{figure}

    To further explore the effects of recombination rate and exon density on the power to detect sweeps, we show in Figure \ref{fig:power-recomb}
    the relationship between power and local recombination rate for our sweep simulations. We see that power to detect sweeps is generally
    a decreasing function of local recombination rate, which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination. We also see that the presence of purifying selection acting on exons has a relatively small effect on the power to detect sweeps,    
    but one that is present across the entire range of recombinations rates observed in human chromosome 1. 
    We also note that the different methods for detecting sweeps that we have explored here respond differently to the effects of recombination rate.
    Unsurprisingly, the empirical cutoff of $\pi$ to detect sweeps is the less robust to the effects of recombination rate than the other two methods,
    that are specifically designed to find selective sweeps. Whereas \sweepfinder2 has a clear decline in power with increasing recombination rate,
    \diploshic 's behavior is more variable, with a slight increase in power at intermediate recombination rates in the CEU sample and a decline
    in power at high recombination rates in the YRI sample.

    In Figure \ref{fig:power-exon} we show the relationship between power and local exon density for our sweep simulations.
    Here there is little relationship between power and exon density, which is somewhat surprising given that the power to detect sweeps
    is generally lower in the presence of purifying selection acting on exons. However while this is the case we see that 
    generally all methods have more power to detect sweeps in regions of lower exon density, which is consistent with the idea that
    background selection acting on nearby exons can reduce the signature of a sweep. However this is far from a simple pattern:
    for instance sweepfinder2 seems to have minimum power in regions of intermediate exon density, while \diploshic is roughly constant
    or perhaps has a slight increase in power at intermediate exon densities. Noise is the estimates of power with respect to exon density
    are prohibitively high for the empirical cutoff of $\pi$ to detect sweeps, so it's not clear what conclusions can be drawn.

@@

>>>>

    In Figure \ref{fig:chr1_power} we show the power to detect selective sweeps at 100 locations along human chromosome 1 from
    the simulations described above. We show the results of three methods for detecting sweeps: 1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos}, and 3) an empirical cutoff of $\pi$ to find regions of reduced diversity.
    A few things are immediately apparent from this figure. First, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that the YRI sample has a larger effective population size and thus
    more variation to detect a sweep. Further the population size of YRI has been relatively constant over the last 100,000 years,
    thus the out-of-Africa bottleneck has does not mute the signal of a sweep as much as in the CEU sample. This is consistent with
    what we know about the effects of demography on the power to detect sweeps \citep[e.g.,][]{simonsen1995properties}.
    Second, the power to detect sweeps is slightly lower in the presence of purifying selection acting on exons but the effect is not
    as strong as one might expect. This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Finally and most importantly, the power to detect sweeps varies considerably among regions of the genome. 
    Indeed variation in power along the chromosome is much greater than the variation in power between the CEU and YRI samples or between
    models with or without purifying selection. This is likely due to the effects of recombination rate heterogeniety as well
    as the density of functional elements along the chromosome.

    Using this set of simulations we can also compare the power of different methods to detect sweeps.
    In general, the power of the methods varies with respect to the population history of the sample.
    For instance \sweepfinder2 has marginally higher power to detect sweeps in the CEU sample than \diploshic,
    likely as a results of the fact that we here trained \diploshic on simulations under a constant size population,
    while the population size history of CEU is characterized by a strong recent bottleneck.
    Conversely, \diploshic has higher power to detect sweeps in the YRI sample than \sweepfinder2,
    likely because that population size history more closely matches the population size history used to train \diploshic.
    We can further explore the performance of the methods by plotting the receiver operating characteristic (ROC) curves for the methods.
    In Figure \ref{fig:roc-curves} we show the ROC curves for the three methods for detecting sweeps.
    False positive rates were established using neutral simulations that contained no sweeps
    but were conditioned on the same population history as the CEU and YRI samples.
    These suggest that the $\pi$ cutoff is a suprisingly effective method for detecting sweeps,
    in these genomic regions at least, but that it lags behind \diploshic and \sweepfinder2 in terms of specificity
    generally. 
    %While this is not exhaustive benchmarking, we aim to give a sense to the reader
    %of how \stdpopsim can be used to benchmark the performance of different methods easily and efficiently.

    \begin{figure}
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/relationship_power_cM.pdf}
        \caption{
        Power to detect selective sweeps at 100 locations along human chromosome 1 under a three population out-of-Africa model \citep{gutenkunst2009inferring}  
        with and without purifying selection as a function of local recombination rate.
        Simulations are as described in Figure \ref{fig:chr1_power}.
        Each panel shows the results of simulations with purifying selection acting on exons (red; labelled BGS)
        and without purifying selection acting on exons (blue; labelled Neutral).
        True positive rate is shown for these methods for samples from the CEU and YRI samples (left and right respectively).
        Points represent the power at each location and trend lines along with 95\% confidence intervals are shown fit to those points.
        }
        \label{fig:power-recomb}
    \end{figure}

    To further explore the effects of recombination rate and exon density on the power to detect sweeps, we show in Figure \ref{fig:power-recomb}
    the relationship between power and local recombination rate for our sweep simulations. We see that power to detect sweeps is
    a decreasing function of local recombination rate for \sweepfinder2 and $pi$ outliers,
    which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination.
    \diploshic on the other hand shows a slight increase in power with recombination rates.
    We also see that the presence of purifying selection acting on exons has a relatively small effect on the power to detect sweeps,    
    but one that is present across the entire range of recombinations rates observed in human chromosome 1. 
    Unsurprisingly, the empirical cutoff of $\pi$ to detect sweeps is the less robust to the effects of recombination rate than the other two methods,
    that are specifically designed to find selective sweeps.

    In Figure \ref{fig:power-exon} we show the relationship between power and local exon density for our sweep simulations.
    Here there is little relationship between power and exon density, which is somewhat surprising given that the power to detect sweeps
    is generally lower in the presence of purifying selection acting on exons. However while this is the case we see that 
    generally all methods have more power to detect sweeps in regions of lower exon density, which is consistent with the idea that
    background selection acting on nearby exons can reduce the signature of a sweep. However this is far from a simple pattern:
    for instance sweepfinder2 seems to have minimum power in regions of intermediate exon density, while \diploshic is roughly constant
    or perhaps has a slight increase in power at intermediate exon densities. Noise in the estimates of power with respect to exon density
    are prohibitively high for the empirical cutoff of $\pi$ to detect sweeps, so it's not clear what conclusions can be drawn.

@@

% roc curves for sweeps
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/roc_neutral_null.pdf}
    \caption{
    \label{fig:roc-curves}
    Receiver operating characteristic (ROC) curves for the three methods for detecting sweeps.
    False positive rates were established using neutral simulations that contained no sweeps
    but were conditioned on the same population history as the CEU and YRI samples.
    }
\end{figure}
